---
output: 
  pdf_document:
    fig_width: 4
    fig_height: 3
header-includes:
  - \pagestyle{empty}
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

This file will give a brief tutorial on how to install and use the ONAM package.  

## Install package:
```{r install onam, eval = FALSE}
install.packages("devtools")  
library(devtools)  
devtools::install_github("Koehlibert/ONAM")
```
If this is the first time using Keras or TensorFlow in R, you need to run `keras::install_keras()`.

The following packages are required for data examples, visualization and machine learning models used for demonstration:

```{r library calls}
library(ONAM)
library(mlbench)
library(MASS)
library(gbm)
library(e1071)
library(caret)
library(xgboost)
library(dplyr)
```

We demonstrate the algorithm by explaining a gradient boosting machine for probability prediction of diabetes, and a xgboost model for prediction of housing prices using the boston housing dataset.

Model specification, evaluation and visualization will be demonstrated.

\newpage

# Binary classification gradient boosting machine for diabetes prediction
The Pima Indians Diabetes Database is a dataset from the Indian National Institute of Diabetes and Digestive and Kidney Dieseases. 
https://www.mdpi.com/2076-3417/9/21/4604

```{r load diabetes data}
data(PimaIndiansDiabetes)
diabetes_data <- PimaIndiansDiabetes %>%
  mutate(diabetes = ifelse(diabetes == "pos", 1, 0)) %>%
  mutate(across(!diabetes, as.numeric)) %>%
  filter(mass > 0)

head(diabetes_data)
```

It consists of `r nrow(diabetes_data)` instances of multiple medical features and the diabetes status of each observation. There are `r sum(diabetes_data$diabetes == 1)` diabetes cases and `r sum(diabetes_data$diabetes == 0)` controls in the dataset.

We demonstrate how to use ONAM to explain the prediction of a gradient boosting machine for prediction of the diabetes status.

First, we fit a gradient boosting machine for prediction of diabtetes probability.

```{r Diabetes gbm, echo = TRUE}
# Train the GBM model (binary classification)
gbm_model <- gbm(diabetes ~ .,
                 data = diabetes_data,
                 distribution = "bernoulli",  
                 n.trees = 500,               
                 interaction.depth = 3,       
                 shrinkage = 0.01,           
                 cv.folds = 5) 
summary(gbm_model, plotit = F)
```

\newpage

Model performance:

```{r gbm performance, echo = FALSE}
predicted_status <- predict(gbm_model, diabetes_data, n.trees = 500)
caret::confusionMatrix(factor(as.numeric(predicted_status > 0.5)), 
                       factor(diabetes_data$diabetes))
```

\newpage

## Fit ONAM model on gbm data

We explain the svm using the orthogonal additive model framework for binary classification:

```{r diabetes onam model, echo = TRUE, results = "hide", eval = FALSE}
Diabetes_formula <- diabetes ~ mod(glucose) + mod(insulin) +
  mod(pregnant) + mod(pressure) + mod(triceps) +
  mod(pedigree) + mod(mass) + mod(age) + 
  mod(glucose, insulin) + mod(age, mass) + mod(.)
list_of_mods_Diabetes <- list(mod = ONAM:::get_submodel)
gbm_expl <- onam(
  Diabetes_formula,
  list_of_deep_models_Diabetes,
  diabetes_data,
  gbm_model,
  prediction_function = function(model, data) {
    predict(model, data, n.trees = 500, type = "response")
  },
  target = "binary",
  n_ensemble = 20,
  progresstext = TRUE,
  epochs = 500,
  verbose = 0
)

#generate prediction object for saving
gbm_res <- predict(gbm_expl)
saveRDS(gbm_res, "gbm_res.RDS")

```

```{r diabetes model summary, echo = FALSE, eval = TRUE}
readRDS("diabetes_summary.rds")
```

```{r gbm diabetes plots, echo = FALSE}
gbm_res <- readRDS("gbm_res.RDS")
```

\newpage

### Example of effect plots

The shown effects are on the logit scale.

A higher plasma glucose concentration is associated with a higher risk of diabetes:

```{r pregnancy effect}
plot_main_effect(gbm_res, "glucose")
```

A higher BMI is associated with a higher risk of diabetes, but the increase is lower in younger people.

```{r mass age effect, eval=FALSE}
plot_main_effect(gbm_res, "mass") + xlab("BMI")

plot_inter_effect(gbm_res, "age", "mass", interpolate = FALSE) +
    xlab("BMI") + ylab("Age")
```

```{r hidden plot library, echo = FALSE}
library(ggpubr)
library(ggplot2)
```


```{r mass age effect combined, echo = FALSE, fig.width=7}
ggarrange(
  plot_main_effect(gbm_res, "mass") + xlab("BMI"),
  plot_inter_effect(gbm_res, "age", "mass",
                    interpolate = FALSE) +
    xlab("BMI") + ylab("Age") +
    theme(legend.position = "right"),
  ncol = 2,
  widths = c(0.35, 0.65)
) 
```


\newpage

# Prediction of Boston housing prices using xgboost

The `r "Boston"` dataset contains data collected by the US Census Service on socioecological features of housing in Boston, MA. It contains 13 features along with the median value of owner-occupied homes, which will be the outcome in this section.

```{r load Boston data}
data("Boston")

head(Boston)
```

The dataset contains information `r nrow(Boston)` towns. We use xgboost to fit a prediction model for median housing prices per town.

```{r fit xgboost model}
train_Boston <- Boston %>% dplyr::select(-medv)
label_Boston <- Boston %>% dplyr::select(medv) %>% unlist()

xgb_train_Boston <- xgb.DMatrix(as.matrix(train_Boston),
                                   label = label_Boston)

xgb_mod <-
  xgboost(
    xgb_train_Boston,
    nrounds = 10000,
    verbose = 0,
    objective = "reg:squarederror",
    eval_metric = "rmse"
  )
summary(xgb_mod)

preds <- predict(xgb_mod, xgb_train_Boston)
pricing_cor <- cor(preds, Boston$medv)
```

Correlation between housing prices and predictions: `r round(pricing_cor, 4)`

## Fit ONAM model on xgboost data

We explain the xgboost using the orthogonal additive model framework:

```{r onam insurance, echo = TRUE, echo = TRUE, results = "hide", eval = FALSE}
Boston_formula <-
  medv ~ mod(crim) + mod(zn) + mod(indus) + mod(chas) +
  mod(nox) + mod(rm) + mod(age) + mod(dis) + mod(rad) +
  mod(tax) + mod(ptratio) + mod(black) + mod(lstat) +
  mod(crim, dis) + mod(rm, age) + mod(zn, indus) +
  mod(.)

list_of_deep_models_Boston <- list(mod = ONAM:::get_submodel)

categorical_features <- c("chas")

xgb_expl <-
  onam(
    Boston_formula,
    list_of_deep_models_Boston,
    Boston,
    # target = "binary",
    model = xgb_mod,
    model_data = xgb_train_Boston,
    categorical_features = categorical_features,
    n_ensemble = 20,
    epochs = 1000,
    progresstext = TRUE,
    verbose = 0
  )

xgb_res <- predict(xgb_expl)
saveRDS(xgb_res, "xgb_res.RDS")
```

```{r xgb summary Boston, echo=FALSE}
readRDS("xgb_summary.rds")
```

```{r load and plot xgb onam model, echo=FALSE}
xgb_res <- readRDS("xgb_res.RDS")
```

### Examples of effect plots

Higher crime rate is associated with lower housing prices

```{r crime effect}
plot_main_effect(xgb_res, "crim") + xlab("Crime rate (%)")
```

Lower number of rooms is associated with lower housing prices, but less so in (areas with) older houses.

```{r room effect, eval = FALSE}
plot_main_effect(xgb_res, "rm") + xlab("Av. number of rooms per dwelling")
plot_inter_effect(xgb_res, "rm", "age") + 
  xlab("Av. number of rooms per dwelling") + 
  ylab("% of units build prior to 1940")
```

```{r room effect combined, echo = FALSE, fig.width=7}
ggarrange(
  plot_main_effect(xgb_res, "rm") +
    xlab("Av. number of rooms per dwelling"),
  plot_inter_effect(xgb_res, "rm", "age") +
    xlab("Av. number of rooms per dwelling") +
    ylab("% of units build prior to 1940") +
    theme(legend.position = "right"),
  ncol = 2,
  widths = c(0.35, 0.65)
)
```

